---
output:
  pdf_document: default
---
## Milestone 2 - Predict Flight Delays
### Tarin Eccleston

## 1. Goal

The main objective of my project is to analyse the 


## 2. Data Source

Talk about process of getting data, issues with getting flight data through APIs
- FAA Data through Kaggle
- Weather data through OpenWeather

```{r}

```

## 3. Data Processing

- 

## 4. Data Exploration

Let's take a quick look at our data...

```{r}
setwd("/Users/tarineccleston/Documents/Software:DS/predict-flight-delays")
library(tidyverse)

flights_df = read.csv("../output/flights.csv")
```

```{r}
glimpse(flights_df)
```

```{r}

```


### 4.1 Question 1: Flight Delays between airports

The original exploration question was to compare flight delays between airports. Are some airports more likely to have flight delays due to a variety of factors which cannot be accounted for in the data, such as operation issues?

Luckily our data gies us the number of minutes delayed due to air system, security, airline, late aircraft and weather. 

- 4.2 Question 2:
- 4.3 Question 3:

## 5. Analytical Plan


## 6. Appendix

## Data cleaning

```{r}

```

### Python Code to Extract Weather Data

```{r}

```

